---
title:
  "NYPD Shooting Incident Data Analysis (Update Version) -
  Temporal pattern in shooting incidents in NYC - 2006-20024"
author:
  "CA"
date:
  "2025-12-02"
output:
  pdf_document
fontsize:
  12pt
---
List of libraries used in this project:
  car | tidyverse | readxl  | broom | reghelper | knitr

# Summary

In this project, I used the NYPD Shooting Incident data of the period 2006-2023 from the NYC Open Data, a free public data repository published by New York City Agencies and other partners. I explored if the victims of shootings in New York during the years between 2006 and 2023 period were mostly female and, if there was a weekly and/or monthly pattern in the occurrence.

# Objective

- To explore if the victims of shootings in New York during the years between 2006 and 2023 period are mostly female 
- To determine if there is a weekly and/or monthly pattern in the occurrence.

# Setup

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

# Set the seed to generate the same random sequence of pseudo-numbers each time that this code is run. This ensures reproducibility.

set.seed(111)

```


## Load libraries

```{r libraries}

# Load libraries

library(car)
library(tidyverse)
library(readxl)
library(broom)
library(reghelper)
library(knitr)
```

## Load data

```{r incident.data-load}

# Create string with the url to download the data with shooting incidents
url.ini.data <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"

# Load data
incident.data <- read_csv(url.ini.data)

# Get a glimpse of the data

glimpse(incident.data)

```

# Features table

I extracted the table with the abbreviations and descriptions of the variables from NYC Open Data website  <https://data.cityofnewyork.us/Public-Safety/NYPD-Shooting-Incident-Data-Historic-/833y-fsy8/about_data> . 

| **Column Name**         | **Column Description**|                             
|-------------------------|------------------------
| INCIDENT_KEY             | Randomly generated persistent ID for each incident |
| OCCUR_DATE               | Exact date of the shooting incident |            
| OCCUR_TIME               | Exact time of the shooting incident |            
| BORO                     | Borough where the shooting incident occurred | 
| PRECINCT                 | Precinct where the shooting incident occurred |  
| JURISDICTION_CODE        | Jurisdiction where the shooting incident | occurred |
| LOCATION_DESC            | Location of the shooting incident|               | STATISTICAL_MURDER_FLAG  | Shooting resulted in the victim's death which would be counted as a murder |
| PERP_AGE_GROUP           | Perpetrator's age within a category |            
| PERP_SEX                 | Perpetrator's sex description |                  
| PERP_RACE                | Perpetrator's race description |                 
| VIC_AGE_GROUP            | Victim's age within a category |                 
| VIC_SEX                  | Victim's sex description |                       
| VIC_RACE                 | Victim's race description |                      
| X_COORD_CD               | Midblock X-coordinate for New York State Plane Coordinate System, Long Island Zone, NAD 83, units feet (FIPS 3104) |         
| Y_COORD_CD               | Midblock Y-coordinate for New York State Plane  Coordinate System, Long Island Zone, NAD 83, units feet (FIPS 3104) |

In the same website, I found that INCIDENT_KEY could appear in multiple rows because there could be more than one victim during a shooting incident. Therefore, each row represents a victim.

# Data Wrangling

For this analysis, I used the following features: 

* OCCUR_DATE
* VIC_SEX 

```{r select}

incident.data <- incident.data %>%
  select(INCIDENT_KEY, OCCUR_DATE, VIC_SEX)

```

## Missing values 

```{r missing}

# Check missing values in INCIDENT_key

sprintf('Missing values for INCIDENT_key: %d',  sum(is.na(incident.data$INCIDENT_KEY)))

# Missing values for OCCUR_DATE
sprintf('Missing values for OCCUR_DATE: %d',  sum(is.na(incident.data$OCCUR_DATE)))

# Missing values for VICTIMS'SEX:
sprintf('Missing values for VIC_SEX: %d',  sum(is.na(incident.data$VIC_SEX)))


```

## Data Transformation

1. I changed the type of data in OCCUR_DATE from 'character' to 'date' type to get year of each entry.

```{r parse_date}

# Transform OCCUR_DATA to POSIXct date-time type.

cleaned.data <- incident.data %>%
  mutate(OCCUR_DATE = as.Date(OCCUR_DATE,
                                 tz = '',
                                 tryFormats = '%m/%d/%Y'))  


# To be able to work with the date per year, a new column 'year' needs to be created.

cleaned.data <- cleaned.data %>%
  mutate(
    # Create a column with year data
    YEAR = format(OCCUR_DATE, '%Y'),
    YEAR = factor(YEAR, levels = sort(unique(YEAR)))
  )

```


2. I converted column VIC_SEX to factor and changed levels name of the sex of victims and reordered the levels.

```{r vic-info}

# Check the labels used for the sex of victims

unique(cleaned.data$VIC_SEX)

# The levels "F", "M" and "U" need to be changed to "Female", "Male" and "Unknown".

cleaned.data <- cleaned.data %>%
  mutate(VIC_SEX = parse_factor(VIC_SEX),
         VIC_SEX = recode(VIC_SEX,
                'F' = 'Female',
                'M' = 'Male',
                'U' = 'Unknown'),
         VIC_SEX = fct_relevel(VIC_SEX,
                     'Unknown',
                     'Female',
                     'Male'))

# Check that the transformation was successful
unique(cleaned.data$VIC_SEX)

```

## Data Visualization : Are the victims of shooting incidents in New York mostly female?

With the next graph I aimed to explore the question if the victims of shooting incidents during period 2006-2013 in New York were mostly women.

```{r graph-victim}

# Plot the number of victims female, male victims by year. The number of victims with unknown sex was not included in this analysis due to his low number.

cleaned.data <- cleaned.data %>%
  filter(VIC_SEX != 'Unknown')

ggplot(data = cleaned.data, aes(x = YEAR, fill = VIC_SEX)) +
  geom_bar() + facet_grid(.~ VIC_SEX) +
  theme(axis.text.x = element_text(angle = 80,
                                   vjust = 0.5,
                                   hjust = 0.5,
                                   size = 12),
        axis.title.x = element_text(size = 13),
        axis.text.y = element_text(size = 12),
        axis.title.y = element_text(size = 13),
        panel.background = element_rect(fill = 'white'),
        panel.grid.major = element_blank(),
        axis.line = element_line(color = 'darkgray'),
        strip.text = element_text(size = 13),
        legend.position = 'none') +
  scale_y_continuous(limits = c(0,2000)) +
  scale_fill_manual(values = c('Female' = 'black', 'Male' = 'burlywood3')) +
  ylab('Number of \nshooting vicims per year') +
  xlab('Year')

  
```

Conclusion 1:  Females were not the most common victims during 2006-2023. The number of male victims is higher than number of female victims across the period 2006-2023.

# Bias Identification: 
The previous graph helped me to identified my bias. I was expecting to find that most of the victims were female and that was why initially I decided to focus in the female population. For this reason, and to have a clearer idea of the pattern of shooting in NYC, I redirected the study to include the analysis of the pattern of shooting involving male victims.
I omitted the analysis of the Unknown class because the number of cases is small.

Note: Other sources of bias could be how I grouped the months.


## Data Visualization 2 : Is there a weekly and/or monthly pattern in the number of shootings incidents that involved male and female victims in NY in the period 2006-2023?

### Data Wrangling


The next graph seeks to explore the number of shootings in the different days of week and months of the year during the period 2006-2023 for female and male victims separately. To facilitate the visualization, days were divided in weekdays and weekends, and months, in seasons.

```{r, week-season}

date.transformation <- function(x){
  x <- x %>%
  mutate(
  # Create a column with the day of the week
    DAY_OF_WEEK = weekdays(OCCUR_DATE),
  # Create a new column for the type of day
    WEEK = case_when(
      DAY_OF_WEEK %in% c('Monday', 'Tuesday', 'Wednesday','Thursday', 'Friday') ~ 'Weekday',
      DAY_OF_WEEK %in% c('Saturday', 'Sunday') ~ 'Weekend'),
  # Create a column with month data
    MONTH = months(OCCUR_DATE),
  #Create a new column for Season and assign season according to the month
    SEASON = MONTH,
    SEASON = case_when(
    SEASON %in% c('December', 'January', 'February') ~ 'Winter',
    SEASON %in% c('March', 'April', 'May') ~ 'Spring',
    SEASON %in% c('June', 'July', 'August') ~ 'Summer',
    SEASON %in% c('September', 'October', 'November') ~ 'Fall' ))
  # Convert season to a factor and change levels order to ease analysis

}



cleaned.data <- date.transformation(cleaned.data) %>%
  # Convert season to a factor and change levels order to ease analysis
  mutate(
      SEASON = fct_relevel(SEASON, 'Winter','Spring', 'Summer', 'Fall'), 
  # Create a column with month data
      WEEK = fct_relevel(WEEK, 'Weekday', 'Weekend'))


```


# Data Visualization

To help with visualization, the total number of victims per date were added and then averaged per type of day (weekend or weekday), season and year.

```{r, female-graph}

# Filter female victims
fema.data <- cleaned.data %>%
  filter(VIC_SEX != 'Male') %>%
  # Group by Date and then add the number of female victims per date
  group_by(OCCUR_DATE, WEEK, SEASON, YEAR)%>%
  summarize(VIC_PER_DAY = n()) %>%
  # Group by Week, YEAR, SEASON
  group_by(WEEK, YEAR, SEASON) %>%
  summarize(VIC_PER_DAY = mean(VIC_PER_DAY))

# Plot female victims per season and day of the week

fema.plot = 
  ggplot(data = fema.data,
       aes(x = SEASON,
           y = VIC_PER_DAY,
           shape = WEEK)) +
  geom_jitter(position = position_jitter(width = 0.15),
              size = 4,
              fill = 'burlywood',
              color = 'black',
              stroke = 1) +
  ylab('Average daily number\n of shooting victims') +
   theme(panel.background = element_rect(fill = 'white'),
        panel.grid.major = element_blank(),
        axis.line = element_line(color = 'darkgray'),
        strip.text = element_text(size = 13),
        legend.position = 'top',
        axis.title.y = element_text(size = 14),
        axis.text.x = element_text(size = 13),
        axis.text.y = element_text(size = 13),
        axis.title.x = element_text(size = 14),
        legend.text = element_text(size = 13),
        legend.title = element_text(size = 14)) +
  scale_shape_manual(values = c(23,17))  +
  ggtitle('Average Daily Number of Females Victims per Type of Day, Year and Season')

```


```{r male-graph}

# Filter male victims
male.data <- cleaned.data %>%
  filter(VIC_SEX != 'Female') %>%
  # Group by Date and then add the number of male victims per date
  group_by(OCCUR_DATE, WEEK, SEASON, YEAR)%>%
  summarize(VIC_PER_DAY = n()) %>%
  # Group by Week, YEAR, SEASON
  group_by(WEEK, YEAR, SEASON) %>%
  summarize(VIC_PER_DAY = mean(VIC_PER_DAY))

# Plot male victims per season and day of the week
male.plot =
  ggplot(data = male.data,
       aes(x = SEASON,
           y = VIC_PER_DAY,
           shape = WEEK)) +
  geom_jitter(position = position_jitter(width = 0.15),
              size = 4,
              fill = 'burlywood',
              color = 'black',
              stroke = 1) +
  ylab('Average daily number\n of shooting victims') +
   theme(panel.background = element_rect(fill = 'white'),
        panel.grid.major = element_blank(),
        axis.line = element_line(color = 'darkgray'),
        strip.text = element_text(size = 13),
        legend.position = 'top',
        axis.title.y = element_text(size = 14),
        axis.text.x = element_text(size = 13),
        axis.text.y = element_text(size = 13),
        axis.title.x = element_text(size = 14),
        legend.text = element_text(size = 13),
        legend.title = element_text(size = 14)) +
  scale_shape_manual(values = c(23,17))  +
  ggtitle('Average Daily Number of Males Victims per Type of Day, Year and Season')

```

```{r, panel-graph}

# Plot female and male plot 

fema.plot
male.plot

```

Conclusion: There were more shootings involving male victims on weekends, and they were higher in Summer during the period 2006-2023. In the case of female victims, there seem to be a trend showing an increase in summer.


## Model


In the this section I tested if in the number of victims, female and male separately, were statistically significant higher in summer months: June, July and August, and if they were higher during the weekends through the years 2006-2023. To simplify the model, I tested weekdays vs weekends, and between group months according to seasons.


# Data Wrangling 

Some days during the period 2006-2023 might have not got registered shooting accidents, or they did have registered  shootings but involved victims who were not assigned a sex (I removed those rows in a previous section). I added these days if they existed.

```{r, missing-days}

# Check for days without records
uniq.days = unique(cleaned.data$OCCUR_DATE)

# Create a sequence from the first day with records in 2006

# and the last day with records in 2023.
all.days = seq.Date(min(as.Date(uniq.days)), max(as.Date(uniq.days)), by = 'day')

# Confirm that there are missing days in the dataset.
length(uniq.days)
length(all.days) 

# Convert sequence of total days during 2006-2023 to date data and get day of week, month and year.
all.days = as.Date(all.days)
all.days = as_tibble(all.days)
names(all.days) = 'OCCUR_DATE'
all.days <- all.days %>%
  mutate(
    DAY_OF_WEEK = weekdays(OCCUR_DATE),
    YEAR = format(OCCUR_DATE, '%Y'),
    MONTH = months(OCCUR_DATE)
  )
  
# Use a full join to add the missing days to the dataset.
cleaned.data = full_join(cleaned.data, all.days, by = c('OCCUR_DATE', 'DAY_OF_WEEK', 'YEAR', 'MONTH'))

  # Create a new column and assign the value 0 if there was not a victim that day, 1 otherwise
cleaned.data$VIC = ifelse(is.na(cleaned.data$INCIDENT_KEY) == TRUE, 0, 1)

```

# Model

To assess if the number of victims was statistically significant higher during weekends during 2006-2023 and if this increase was season dependent, I used a linear model followed by Two Way Analysis of Variance (ANOVA) with interaction (I considered interaction between seasons and weekends).

Model for female victims:

```{r, model-female}

# Filter out data of male victims, this will keep row with days with female victims and days without victims
fema.data <- cleaned.data %>%
  filter(VIC_SEX != 'Male') %>%
# Group by Date and then add the number of female victim per day 
  group_by(OCCUR_DATE, DAY_OF_WEEK, MONTH, YEAR)%>%
  summarize(VIC_PER_DAY = sum(VIC))

fema.data <- date.transformation(fema.data) %>%
# Convert SEASON and WEEK to factor
mutate(
  SEASON = factor(SEASON),
  WEEK = factor(WEEK)) %>%
# Group by Week, YEAR, SEASON
  group_by(WEEK, YEAR, SEASON) %>%
  summarize(VIC_PER_DAY = mean(VIC_PER_DAY))

# Linear Regression model with interaction
model.fema.aov <- fema.data %>%
  lm(VIC_PER_DAY ~ SEASON +
                   WEEK + 
                   WEEK : SEASON,
                   data = .,
                   contrasts = list(
                     SEASON = contr.sum,
                     WEEK = contr.sum)
     ) %>% 
  Anova(type = 3)

# Print result  
kable(model.fema.aov, caption = 'Anova test of the Linear Regression Model with Interaction for Female Victims Data')

# The effect of season and the interaction term were not significant, so I removed them from the model.

# New model
model.fema <- fema.data%>%
  lm(VIC_PER_DAY ~ WEEK,
                   data = .) %>%
  tidy(conf.int = TRUE)

# Print results
kable(model.fema, caption = 'Linear Regression Model for Female Victims Data')

```


Model for male victims:

```{r, model-male}

# Filter out data of male victims, this will keep row with days with male victims and days without victims
male.data <- cleaned.data %>%
  filter(VIC_SEX != 'Female') %>%
# Group by Date and then add the number of female victim per day 
  group_by(OCCUR_DATE, DAY_OF_WEEK, MONTH, YEAR)%>%
  summarize(VIC_PER_DAY = sum(VIC))

male.data <- date.transformation(male.data) %>%
# Convert SEASON and WEEK to factor
mutate(
  SEASON = factor(SEASON),
  WEEK = factor(WEEK)) %>%
# Group by Week, YEAR, SEASON
  group_by(WEEK, YEAR, SEASON) %>%
  summarize(VIC_PER_DAY = mean(VIC_PER_DAY))

# Linear Regression model with interaction
model.male.aov <- male.data %>%
  lm(VIC_PER_DAY ~ SEASON +
                   WEEK + 
                   WEEK : SEASON,
                   data = .,
                   contrasts = list(
                     SEASON = contr.sum,
                     WEEK = contr.sum)
     ) %>% 
  Anova(type = 3)

# Print result  
kable(model.male.aov, caption = 'Anova test of the linear regression model with interaction for Male Victims Data')

# The effect of week, season and the interaction term were significant in this case, so I kept them in the model.

# New model
model.male <- male.data%>%
  lm(VIC_PER_DAY ~ SEASON +
                   WEEK +
                   SEASON : WEEK,
                   data = .) %>%
  tidy(conf.int = TRUE)

# Print results
kable(model.male, caption = 'Linear Regression Model for Male Victims Data')

```

# Final Results of Modelling

```{r, model-final-results}

# Print result with the significant terms

kable(model.fema[!(model.fema$conf.low < 0 & model.fema$conf.high > 0), ], caption = 'RESULTS: Table of significant terms of a linear regression model of the number of female victims of shooting per day during 2006-2003')

kable(model.male[!(model.male$conf.low < 0 & model.male$conf.high > 0),], caption = 'RESULTS: Table of significant terms of a linear regression model of the number of male victims of shooting per day during 2006-2003')

```


# Conclusions

In the case of females, I found that there was a small but statistically significant increase on the number of female victims per day during the weekends compared to the rest of the days of the weeks during 2006-2024, but this increase was not affected by seasonality. 
However, in the case of males, I found that there was a statistically significant increase on the number of male victims per day during the weekend compared to the rest of the days of the week during 2006-2024 and this increase was statistically significant higher during fall and summer. The latter was the season where the rate of increased peaked.

# R session information 

```{r session-info}
sessionInfo()

```

